Minimal trap spaces of Boolean networks are maximal siphons of their Petri net encoding

### 2022-11-14 manuscripts/TCS2022/tcs2022.tex
manuscripts

\documentclass[runningheads]{llncs}
\usepackage[T1]{fontenc}
\usepackage{hyperref}
\usepackage{xcolor}
\usepackage{booktabs}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Drawing
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\usepackage{tikz}
\usepackage{amsfonts}

@@

% hyperref for LNCS
\renewcommand\UrlFont{\color{blue}\rmfamily}

@@

@@

\title{Minimal trap spaces of Logical models are maximal siphons of their Petri net encoding}
\titlerunning{Minimal trap spaces as maximal siphons}

\author{Van-Giang Trinh\inst{1}\orcidID{0000-0001-6581-998X} \and \\
  Belaid Benhamou\inst{1} \and \\
  Kunihiko Hiraishi\inst{2}\orcidID{0000-0003-1750-1891} \and \\
  Sylvain Soliman\inst{3}\orcidID{0000-0001-5525-7418}
  }

\authorrunning{Trinh et al.}

\institute{%
  LIS, Aix-Marseille Université, Marseille, France\\
  \email{\{trinh.van-giang, belaid.benhamou\}@lis-lab.fr}
  \and
  School of Information Science, Japan Advanced Institute of Science and Technology, Japan\\
  \email{hira@jaist.ac.jp}
  \and
  Lifeware team, Inria Saclay center, Palaiseau, France\\
  \email{Sylvain.Soliman@inria.fr}
}

\maketitle

\begin{abstract}

  Boolean modelling of gene regulation but also of post-trans\-criptomic systems has proven over the years that it can bring powerful analyses and corresponding insight to the many cases where precise biological data is not sufficiently available to build a detailed quantitative model.
  This is even more true for very large models where such data is frequently missing and led to a constant increase in size of logical models \emph{à la} Thomas.
  Besides simulation, the analysis of such models is mostly based on attractor computation, since those correspond roughly to observable biological \emph{phenotypes}. The recent use of trap spaces made a real breakthrough in that field allowing to consider medium-sized models that used to be out of reach.
  However, with the continuing increase in model-size, the state-of-the-art computation of minimal trap spaces based on \emph{prime-implicants} shows its limits as there can be a huge number of implicants.

  In this article we present an alternative method to compute minimal trap spaces, and hence complex attractors, of a Boolean model. It replaces the need for prime-implicants by a completely different technique, namely the enumeration of maximal siphons in the Petri net encoding of the original model.
  After some technical preliminaries, we expose the concrete need for such a method and detail its implementation using Answer Set Programming.
  We then demonstrate its efficiency and compare it to implicant-based methods on some large Boolean models from the literature.

\keywords{Logical models \and Boolean models \and Trap spaces \and Attractor computation \and Petri nets \and Siphons}
\end{abstract}

\section{Introduction}

@@

Boolean modelling has proven over the years that it can bring powerful analyses and corresponding insight to the many cases where precise biological data is not sufficiently available to build a detailed quantitative model~\cite{wang2012boolean}, even for modelling post-transcriptional mechanisms.
This is even more true for very large models where such data is frequently missing and led to a constant increase in size of logical models \emph{à la} Thomas~\cite{aghamiri2020automated}.
Besides simulation, the analysis of such models is mostly based on attractor computation, since those correspond roughly to observable biological \emph{phenotypes}.
The recent use of trap spaces~\cite{klarner2015computing} made a real breakthrough in that field allowing to consider medium-sized models that used to be out of reach.
However, with the continuing increase in model-size, the state-of-the-art computation of minimal trap spaces based on \emph{prime-implicants} shows its limits as there can be a huge number of implicants.

It is worth noting that the recent method presented in~\cite{DBLP:conf/ictai/ChevalierFPZ19} for computing minimal trap spaces avoids the prime-implicants computation by relying on the \emph{most-permissive} semantics of Boolean models.
This method has been implemented in the tool mpbn\footnote{\url{https://github.com/bnediction/mpbn}} demonstrated in~\cite{Paulev2020} for handling medium-sized models from the literature and very large synthetic models (up to 100,000 nodes).
However, this method is only applicable for \emph{locally-monotonic} Boolean models, whereas the prime-implicants based method~\cite{klarner2015computing} is applicable for \emph{general} Boolean models (i.e., including both locally-monotonic and non-locally-monotonic ones).
The study~\cite{DBLP:journals/tcs/NoualRS13} highlights the need for non-locally-monotonic Boolean models in both biological and theoretical aspects.
Hence, it is still necessary to develop efficient methods for computing minimal trap spaces of large-scale general Boolean models.

@@

In this article we present an alternative method to compute minimal trap spaces, and hence complex attractors, of a general Boolean model. It replaces the need for prime-implicants by a completely different technique, namely the enumeration of maximal siphons in the Petri net encoding of the original model.
After some technical preliminaries, we expose the concrete need for such a method and detail its implementation using Answer Set Programming.
We then demonstrate its efficiency and compare it to implicant-based methods on several large Boolean models from the literature.

@@

\section{Preliminaries}

We will briefly recall here some preliminaries on Boolean models related to trap spaces and Petri nets.
In the case of multi-level Logical models, an encoding into a Boolean model is always possible~\cite{Didier2011}.

@@

We recall here some definitions from~\cite{klarner2015computing} for the introduction of \emph{trap spaces}.
Minimal trap spaces prove to be a very good approximation of the attractors of a Boolean model under asynchronous update schemes and have become the \emph{de facto} standard way to analyze models of a few tens of \emph{genes}~\cite{klarner2017pyboolnet,cifuentes2020control}.

Given a Boolean model \(\mathcal{M} = (V, F)\) with nodes \(V=(v_{1},\dots,v_{n})\) and Boolean functions \(F=(f_{1},\dots,f_{n})\), its state-space is \(\mathcal{S}_{\mathcal{M}} =\mathbb{B}^{n}\) with \(\mathbb{B} = \{0, 1\}\). A state \(s \in \mathbb{B}^{n}\) is a mapping \(s \colon V \mapsto \mathbb{B}\) that assigns either 0 (inactive) or 1 (active) to each node. At each time step \(t\), node \(v_i\) can update its state by
\[v_i(t + 1) \in \{v_{i}(t), f_i(v(t))\}\]
where \(v(t)\) is the state of \(\mathcal{M}\) at time \(t\), \(v_i(t + 1)\) is the state of node \(v_i\) at time \(t + 1\) and an update scheme specifies which value is chosen~\cite{thomas1991regulatory}.
There are two main types of update schemes: synchronous, where for all \(i, v_{i}(t+1) = f_{i}(v(t))\), and asynchronous, where an \(i\) is chosen among the possible updates (\(f_{i}(v(t))\neq v_{i}(t)\)), and all the other nodes \(j\) remain fixed at \(v_{j}(t)\).

@@

A \emph{trap space} is defined as a subspace that is also a trap set.
It is noted that trap spaces of a Boolean model are independent of the update scheme of this model~\cite{klarner2015computing}.
Then, we define a partial order \(<\) on \(\mathcal{S}_{\mathcal{M}}^{\star}\) as: \(m < m'\) if and only if \(\mathcal{S}_{\mathcal{M}}[m] \subseteq \mathcal{S}_{\mathcal{M}}[m']\) and \(\mathcal{S}_{\mathcal{M}}[m] \neq \mathcal{S}_{\mathcal{M}}[m']\).
Consequently, a trap space \(m\) is minimal if and only if there is no trap space \(m' \in \mathcal{S}_{\mathcal{M}}^{\star}\) such that \(m' < m\).

For example, let us consider the Boolean model shown in Example~\ref{example:BN}.
Figure~\ref{fig:stg} shows the dynamics of this model under the fully asynchronous update (i.e., only one node is nondeterministically selected in order to be updated at each time step).
The model has all two trap spaces, \(m_1 = 11\) and \(m_2 = \star\star\).
Since \(m_1 < m_2\), \(m_1\) is a minimal trap space of the Boolean model.

\begin{example}
We give a Boolean model \(\mathcal{M} = (V, F)\), where \(V = (x_1, x_2)\) and \(F = (f_1, f_2)\) with \(f_1 = (x_1 \land x_2) \lor (\neg x_1 \land \neg x_2), f_2 = (x_1 \land x_2) \lor (\neg x_1 \land \neg x_2)\). Herein, \(\land\), \(\lor\), and \(\neg\) denote the conjunction, disjunction, and negation logical operators, respectively.\label{example:BN}
\end{example}

@@

  \draw[->] (nx2) -- (t1x2);
  \draw[->] (t1x2) -- (x2);
  \draw[<->] (t1x2) -- (nx1);
\end{tikzpicture}
\caption{Petri net encoding of the model. Circles denote places, whereas rectangles denote transitions.}%
\label{fig:PN}
\end{subfigure}%
\caption{Dynamics and encoding of the Boolean model of Example~\ref{example:BN}}%
\label{fig:stg_and_PN}
\end{figure}

\subsection{Petri net encoding of Boolean models}%
\label{sec:encoding}

@@

The link between Boolean models \emph{à la} Thomas and Petri nets was originally established in~\cite{chaouiya2004qualitative} in order to make available formal methods like model-checking for the analysis of such systems.
The basic encoding into 1-safe (i.e., never more than one token in each place) nets only holds for purely Boolean models but was later extended to multivalued Logical models in two ways, either in~\cite{chaouiya2011petri} with non 1-safe Petri nets or more recently in~\cite{chatain2014characterization} with 1-safe nets but many more places.

Since our study is focused on Boolean models, we briefly recall the original encoding here.
Its basis is that every node (\emph{gene}) \(v\) of the original model \(\mathcal{M} = (V, F)\) is represented by two separate places (\(p_v\) and \(\overline{p}_v\)), corresponding to its two states, active, and inactive, respectively.
Each conjunct of the logical function that activates the \emph{gene} will lead to a transition \(t\), consuming the inactive place (i.e., a directional arc from \(\overline{p}_v\) to \(t\)), producing the active place (i.e., a directional arc from \(t\) to \(p_v\)), and with all other literals both consumed and produced (i.e., a bidirectional arc).
And conversely for the inactivation.
Let \(s\) be a state of the Boolean model and \(m_s\) be its corresponding marking in the encoded Petri net. It holds that \(\forall v \in V\), \(s(v) = 0\) if and only if \(m_s(\overline{p}_v) = 1\) and \(s(v) = 1\) if and only if \(m_s(p_v) = 1\). Note also that at any marking \(m\) of the Petri net encoding a Boolean model, it always holds that \(m(p_v) + m(\overline{p}_v) = 1\).

The main property of this encoding is that it is completely faithful with respect to the update scheme of the original Boolean model.
For each node \(v\) of \(\mathcal{M}\), only transitions corresponding to \(v\) can change the current marking of \(p_v\) or \(\overline{p_v}\).
In addition, at any marking at most one of such transitions is enabled because \(m(p_v) + m(\overline{p}_v) = 1\) holds.
Hence, for any update scheme in \(\mathcal{M}\), we have a corresponding firing scheme in \(\mathcal{P}\), which preserves the equivalence between the dynamics of \(\mathcal{M}\) and \(\mathcal{P}\)~\cite{DBLP:journals/nc/ChatainHKPT20}.

For illustration, let us reconsider the Boolean model shown in Example~\ref{example:BN}.
Figure~\ref{fig:PN} shows the Petri net encoding of this Boolean model.
Place \(p_{x_1}\) (resp. \(\overline{p_{x_1}}\)) in \(\mathcal{P}\) represents the activation (resp.\ the inactivation) of node \(x_1\) in \(\mathcal{M}\).
Marking \(\{p_{x_1}, \overline{p_{x_2}}\}\) in \(\mathcal{P}\) represents state 10 in \(\mathcal{M}\).
Transitions \(t^{1}_{x_1}\) and \(t^{2}_{x_1}\) represent the update of node \(x_1\).
Of course, in any marking \(t^{1}_{x_1}\) and \(t^{2}_{x_1}\) cannot be both enabled.
Then, the fully asynchronous update scheme in \(\mathcal{M}\) corresponds to the classical firing scheme in \(\mathcal{P}\) where only one of the enabled transitions for a given marking will be fired~\cite{Murata1989}.

Note that given a Boolean model in the standard SBML-Qual format~\cite{chaouiya2013sbml}, i.e., the package of SBML v3~\cite{keating2020sbml} for such models, one can easily obtain its Petri net encoding in the Petri Net Markup Language  (PNML)\footnote{\url{https://www.pnml.org/}} standard using the BioLQM\footnote{\url{http://www.colomoto.org/biolqm/}} library.
This piece of software extracted from GINsim~\cite{chaouiya2012logical} and part of the CoLoMoTo\footnote{\url{http://colomoto.org/}}~\cite{naldi2015cooperative} software suite allows for easy conversion between standard formats.
It also accepts many other common formats for Boolean models, notably the \verb|.bnet| files of the  BoolNet~\cite{mussel2010boolnet,klarner2017pyboolnet} tools.
The conversion is executed as follows:

@@

Note that transforming a Boolean model defined by its functions into its Petri net encoding roughly relies on obtaining conditions for the activation and inactivation of the states. In~\cite{chaouiya2004qualitative} this took the form of the whole truth table of the Boolean functions, but as shown in Appendix 1 of~\cite{chatain2014characterization} computing Disjunctive Normal Forms (DNF) of each Boolean function is enough.
Though this might appear quite computationally intensive it is important to remark first that contrary to the prime-implicants case, there is no need to find \emph{minimal} DNFs.
One way to look at this is to consider that this amounts to a similar approach as that used in~\cite{DBLP:conf/ictai/ChevalierFPZ19} but with the encoding of both activation and inhibition functions as DNFs in order to take into account possible non-local-monotonicity.
This does not change the worst-case-complexity (obtaining a single DNF being exponential) but might matter a lot in practice.
As such, we will explore how this transformation, here using BDDs in BioLQM, and the one based on the most-permissive semantics compare in the Section~\ref{sec:eval} on evaluation.

@@

\section{Minimal trap spaces as maximal conflict-free siphons}
First, we add a definition related to any set of places of a Petri net encoding a Boolean model, and notably a siphon of such a net.

@@

  A set of places of Petri net \(\mathcal{P}\) encoding Boolean model \(\mathcal{M}\) is \emph{conflict-free} if it does not contain any two places corresponding to the active and inactive states of the same \emph{gene} of \(\mathcal{M}\). Then, a conflict-free siphon \(S\) is said to be \emph{maximal} if and only if there is no other conflict-free siphon \(S'\) such that \(S \subset S'\).

@@

Intuitively, a siphon is a set of places that once unmarked remains so.
If it is conflict-free then its dual corresponds to a partial-state of the model such that whatever update, the fixed values remain so (since the unmarked places remain unmarked).
This is precisely the definition of a trap space and maximality of the siphon is equivalent to as many fixed values as possible, hence minimality of the trap space.
For example, the Boolean model given in Example~\ref{example:BN} has two trap spaces, \(m_1 = 11\) and \(m_2 = \star\star\).
The Petri net encoding of this Boolean model has five generic siphons, \(S_1 = \emptyset\), \(S_2 = \{p_{x_1}, \overline{p_{x_1}}\}\), \(S_3 = \{p_{x_2}, \overline{p_{x_2}}\}\), \(S_4 = \{\overline{p_{x_1}}, \overline{p_{x_2}}\}\), and \(S_5 = \{p_{x_1}, \overline{p_{x_1}}, p_{x_2}, \overline{p_{x_2}}\}\).
However, only \(S_1\) and \(S_4\) are conflict-free siphons and correspond to \(m_2\) and \(m_1\), respectively.
Since \(S_1 \subset S_4\), \(S_4\) is a maximal siphon corresponding to the minimal trap space \(m_1\).
Hereafter, we formally prove that a maximal conflict-free siphon is equivalent to a minimal trap space.

@@

>>>>

%% 
%% Copyright 2007-2020 Elsevier Ltd
%% 
%% This file is part of the 'Elsarticle Bundle'.
%% ---------------------------------------------
%% 
%% It may be distributed under the conditions of the LaTeX Project Public
%% License, either version 1.2 of this license or (at your option) any
%% later version.  The latest version of this license is in
%%    http://www.latex-project.org/lppl.txt
%% and version 1.2 or later is part of all distributions of LaTeX
%% version 1999/12/01 or later.
%% 
%% The list of all files belonging to the 'Elsarticle Bundle' is
%% given in the file `manifest.txt'.
%% 
%% Template article for Elsevier's document class `elsarticle'
%% with harvard style bibliographic references

\documentclass[preprint,12pt]{elsarticle}

%% Use the option review to obtain double line spacing
%% \documentclass[preprint,review,12pt]{elsarticle}

%% Use the options 1p,twocolumn; 3p; 3p,twocolumn; 5p; or 5p,twocolumn
%% for a journal layout:
%% \documentclass[final,1p,times]{elsarticle}
%% \documentclass[final,1p,times,twocolumn]{elsarticle}
%% \documentclass[final,3p,times]{elsarticle}
%% \documentclass[final,3p,times,twocolumn]{elsarticle}
%% \documentclass[final,5p,times]{elsarticle}
%% \documentclass[final,5p,times,twocolumn]{elsarticle}

%% For including figures, graphicx.sty has been loaded in
%% elsarticle.cls. If you prefer to use the old commands
%% please give \usepackage{epsfig}

%% The amssymb package provides various useful mathematical symbols
\usepackage{amssymb}
%% The amsthm package provides extended theorem environments
\usepackage{amsthm}

\newtheorem{theorem}{Theorem}[section]
\newtheorem{example}{Example}[section]
\newtheorem{definition}{Definition}[section]

%% The lineno packages adds line numbers. Start line numbering with
%% \begin{linenumbers}, end it with \end{linenumbers}. Or switch it on
%% for the whole article with \linenumbers.
\usepackage{lineno}

\usepackage{hyperref}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Drawing
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\usepackage{tikz}
\usepackage{amsfonts}

@@

\usepackage{booktabs}

@@

\journal{Theoretical Computer Science}

@@

\begin{frontmatter}

%% Title, authors and addresses

%% use the tnoteref command within \title for footnotes;
%% use the tnotetext command for theassociated footnote;
%% use the fnref command within \author or \address for footnotes;
%% use the fntext command for theassociated footnote;
%% use the corref command within \author for corresponding author footnotes;
%% use the cortext command for theassociated footnote;
%% use the ead command for the email address,
%% and the form \ead[url] for the home page:
%% \title{Title\tnoteref{label1}}
%% \tnotetext[label1]{}
%% \author{Name\corref{cor1}\fnref{label2}}
%% \ead{email address}
%% \ead[url]{home page}
%% \fntext[label2]{}
%% \cortext[cor1]{}
%% \affiliation{organization={},
%%             addressline={},
%%             city={},
%%             postcode={},
%%             state={},
%%             country={}}
%% \fntext[label3]{}

\title{Minimal trap spaces of Boolean networks are maximal siphons of their Petri net encoding}

%% use optional labels to link authors explicitly to addresses:
%% \author[label1,label2]{}
%% \affiliation[label1]{organization={},
%%             addressline={},
%%             city={},
%%             postcode={},
%%             state={},
%%             country={}}
%%
%% \affiliation[label2]{organization={},
%%             addressline={},
%%             city={},
%%             postcode={},
%%             state={},
%%             country={}}

\author{Van-Giang Trinh\fnref{1}}
\ead{trinh.van-giang@lis-lab.fr}

\author{Belaid Benhamou\fnref{1}}
\ead{belaid.benhamou@lis-lab.fr}

\affiliation[1]{organization={LIS, Aix-Marseille University},%Department and Organization
            %addressline={}, 
            city={Marseille},
            %postcode={}, 
            %state={},
            country={France}}

\author{Sylvain Soliman\corref{cor1}\fnref{2}}
\ead{Sylvain.Soliman@inria.fr}

\affiliation[2]{organization={Lifeware team, Inria Saclay center},%Department and Organization
            %addressline={}, 
            city={Palaiseau},
            %postcode={}, 
            %state={},
            country={France}}

\cortext[cor1]{Corresponding author.}

\begin{abstract}
%% Text of abstract

Boolean networkling of gene regulation but also of post-trans\-criptomic systems has proven over the years that it can bring powerful analyses and corresponding insight to the many cases where precise biological data is not sufficiently available to build a detailed quantitative model.
This is even more true for very large models where such data is frequently missing and led to a constant increase in size of logical models \emph{à la} Thomas.
Besides simulation, the analysis of such models is mostly based on attractor computation, since those correspond roughly to observable biological \emph{phenotypes}. The recent use of trap spaces made a real breakthrough in that field allowing to consider medium-sized models that used to be out of reach.
However, with the continuing increase in model-size, the state-of-the-art computation of minimal trap spaces based on \emph{prime-implicants} shows its limits as there can be a huge number of implicants.

In this article we present an alternative method to compute minimal trap spaces, and hence complex attractors, of a Boolean network. It replaces the need for prime-implicants by a completely different technique, namely the enumeration of maximal siphons in the Petri net encoding of the original model.
After some technical preliminaries, we expose the concrete need for such a method and detail its implementation using Answer Set Programming.
We then demonstrate its efficiency and compare it to implicant-based methods on some large Boolean networks from the literature.

\end{abstract}

%%Graphical abstract
% \begin{graphicalabstract}
% %\includegraphics{grabs}
% \end{graphicalabstract}

%%Research highlights
% \begin{highlights}
% \item Research highlight 1
% \item Research highlight 2
% \end{highlights}

\begin{keyword}
%% keywords here, in the form: keyword \sep keyword

%% PACS codes here, in the form: \PACS code \sep code

%% MSC codes here, in the form: \MSC code \sep code
%% or \MSC[2008] code \sep code (2000 is the default)

Logical model \sep Boolean network \sep Trap space \sep Attractor computation \sep Petri net \sep Siphon \sep Answer set programming \sep Systems biology
\end{keyword}

\end{frontmatter}

\linenumbers

%% main text
\section{Introduction}

@@

Boolean networkling has proven over the years that it can bring powerful analyses and corresponding insight to the many cases where precise biological data is not sufficiently available to build a detailed quantitative model~\cite{wang2012boolean}, even for modelling post-transcriptional mechanisms.
This is even more true for very large models where such data is frequently missing and led to a constant increase in size of logical models \emph{à la} Thomas~\cite{aghamiri2020automated}.
Besides simulation, the analysis of such models is mostly based on attractor computation, since those correspond roughly to observable biological \emph{phenotypes}.
The recent use of trap spaces~\cite{klarner2015computing} made a real breakthrough in that field allowing to consider medium-sized models that used to be out of reach.
However, with the continuing increase in model-size, the state-of-the-art computation of minimal trap spaces based on \emph{prime-implicants} shows its limits as there can be a huge number of implicants.

It is worth noting that the recent method presented in~\cite{DBLP:conf/ictai/ChevalierFPZ19} for computing minimal trap spaces avoids the prime-implicants computation by relying on the \emph{most-permissive} semantics of Boolean networks.
This method has been implemented in the tool mpbn\footnote{\url{https://github.com/bnediction/mpbn}} demonstrated in~\cite{Paulev2020} for handling medium-sized models from the literature and very large synthetic models (up to 100,000 nodes).
However, this method is only applicable for \emph{locally-monotonic} Boolean networks, whereas the prime-implicants based method~\cite{klarner2015computing} is applicable for \emph{general} Boolean networks (i.e., including both locally-monotonic and non-locally-monotonic ones).
The study~\cite{DBLP:journals/tcs/NoualRS13} highlights the need for non-locally-monotonic Boolean networks in both biological and theoretical aspects.
Hence, it is still necessary to develop efficient methods for computing minimal trap spaces of large-scale general Boolean networks.

@@

In this article we present an alternative method to compute minimal trap spaces, and hence complex attractors, of a general Boolean network. It replaces the need for prime-implicants by a completely different technique, namely the enumeration of maximal siphons in the Petri net encoding of the original model (Section~\ref{sec:Main_finding}).
After some technical preliminaries (Section~\ref{sec:Preliminaries}), we expose the concrete need for such a method and detail its implementation using Answer Set Programming~\ref{sec:Computation}.
We then demonstrate its efficiency and compare it to implicant-based methods on several large Boolean networks from the literature (Section~\ref{sec:eval}).

@@

% Mention the extensions for the journal version
Herein we extend our previous work in~\cite{DBLP:conf/cmsb/TrinhBHS22} as follows.
First, beyond the proposed ASP method, we propose several other possible methods using MaxSAT, CLP, and ILP.
Second, we conduct a more comprehensive benchmark to evaluate all the proposed methods.
Third, regarding the implementation, we have developed a converter that directly reads a .bnet file and builds the Petri net encoding, instead of using the PNML conversion of BioLQM.
Finally, we explore several complexity analysis on maximal (conflict-free) siphon problems.

\section{Preliminaries}
\label{sec:Preliminaries}

We will briefly recall here some preliminaries on Boolean networks related to trap spaces and Petri nets.
In the case of multi-level Logical models, an encoding into a Boolean network is always possible~\cite{Didier2011}.

@@

% TODO give formal definitions

We recall here some definitions from~\cite{klarner2015computing} for the introduction of \emph{trap spaces}.
Minimal trap spaces prove to be a very good approximation of the attractors of a Boolean network under asynchronous update schemes and have become the \emph{de facto} standard way to analyze models of a few tens of \emph{genes}~\cite{klarner2017pyboolnet,cifuentes2020control}.

Given a Boolean network \(\mathcal{M} = (V, F)\) with nodes \(V=(v_{1},\dots,v_{n})\) and Boolean functions \(F=(f_{1},\dots,f_{n})\), its state-space is \(\mathcal{S}_{\mathcal{M}} =\mathbb{B}^{n}\) with \(\mathbb{B} = \{0, 1\}\). A state \(s \in \mathbb{B}^{n}\) is a mapping \(s \colon V \mapsto \mathbb{B}\) that assigns either 0 (inactive) or 1 (active) to each node. At each time step \(t\), node \(v_i\) can update its state by
\[v_i(t + 1) \in \{v_{i}(t), f_i(v(t))\}\]
where \(v(t)\) is the state of \(\mathcal{M}\) at time \(t\), \(v_i(t + 1)\) is the state of node \(v_i\) at time \(t + 1\) and an update scheme specifies which value is chosen~\cite{thomas1991regulatory}.
There are two main types of update schemes: synchronous, where for all \(i, v_{i}(t+1) = f_{i}(v(t))\), and asynchronous, where an \(i\) is chosen among the possible updates (\(f_{i}(v(t))\neq v_{i}(t)\)), and all the other nodes \(j\) remain fixed at \(v_{j}(t)\).

@@

A \emph{trap space} is defined as a subspace that is also a trap set.
It is noted that trap spaces of a Boolean network are independent of the update scheme of this model~\cite{klarner2015computing}.
Then, we define a partial order \(<\) on \(\mathcal{S}_{\mathcal{M}}^{\star}\) as: \(m < m'\) if and only if \(\mathcal{S}_{\mathcal{M}}[m] \subseteq \mathcal{S}_{\mathcal{M}}[m']\) and \(\mathcal{S}_{\mathcal{M}}[m] \neq \mathcal{S}_{\mathcal{M}}[m']\).
Consequently, a trap space \(m\) is minimal if and only if there is no trap space \(m' \in \mathcal{S}_{\mathcal{M}}^{\star}\) such that \(m' < m\).

For example, let us consider the Boolean network shown in Example~\ref{example:BN}.
Figure~\ref{fig:stg} shows the dynamics of this model under the fully asynchronous update (i.e., only one node is nondeterministically selected in order to be updated at each time step).
The model has all two trap spaces, \(m_1 = 11\) and \(m_2 = \star\star\).
Since \(m_1 < m_2\), \(m_1\) is a minimal trap space of the Boolean network.

\begin{example}
We give a Boolean network \(\mathcal{M} = (V, F)\), where \(V = (x_1, x_2)\) and \(F = (f_1, f_2)\) with \(f_1 = (x_1 \land x_2) \lor (\neg x_1 \land \neg x_2), f_2 = (x_1 \land x_2) \lor (\neg x_1 \land \neg x_2)\). Herein, \(\land\), \(\lor\), and \(\neg\) denote the conjunction, disjunction, and negation logical operators, respectively.\label{example:BN}
\end{example}

@@

  \draw[->] (nx2) -- (t1x2);
  \draw[->] (t1x2) -- (x2);
  \draw[<->] (t1x2) -- (nx1);
\end{tikzpicture}
\caption{Petri net encoding of the model. Circles denote places, whereas rectangles denote transitions.}%
\label{fig:PN}
\end{subfigure}%
\caption{Dynamics and encoding of the Boolean network of Example~\ref{example:BN}}%
\label{fig:stg_and_PN}
\end{figure}

\subsection{Petri net encoding of Boolean networks}%
\label{sec:encoding}

@@

The link between Boolean networks \emph{à la} Thomas and Petri nets was originally established in~\cite{chaouiya2004qualitative} in order to make available formal methods like model-checking for the analysis of such systems.
The basic encoding into 1-safe (i.e., never more than one token in each place) nets only holds for purely Boolean networks but was later extended to multivalued Logical models in two ways, either in~\cite{chaouiya2011petri} with non 1-safe Petri nets or more recently in~\cite{chatain2014characterization} with 1-safe nets but many more places.

Since our study is focused on Boolean networks, we briefly recall the original encoding here.
Its basis is that every node (\emph{gene}) \(v\) of the original model \(\mathcal{M} = (V, F)\) is represented by two separate places (\(p_v\) and \(\overline{p}_v\)), corresponding to its two states, active, and inactive, respectively.
Each conjunct of the logical function that activates the \emph{gene} will lead to a transition \(t\), consuming the inactive place (i.e., a directional arc from \(\overline{p}_v\) to \(t\)), producing the active place (i.e., a directional arc from \(t\) to \(p_v\)), and with all other literals both consumed and produced (i.e., a bidirectional arc).
And conversely for the inactivation.
Let \(s\) be a state of the Boolean network and \(m_s\) be its corresponding marking in the encoded Petri net. It holds that \(\forall v \in V\), \(s(v) = 0\) if and only if \(m_s(\overline{p}_v) = 1\) and \(s(v) = 1\) if and only if \(m_s(p_v) = 1\). Note also that at any marking \(m\) of the Petri net encoding a Boolean network, it always holds that \(m(p_v) + m(\overline{p}_v) = 1\).

The main property of this encoding is that it is completely faithful with respect to the update scheme of the original Boolean network.
For each node \(v\) of \(\mathcal{M}\), only transitions corresponding to \(v\) can change the current marking of \(p_v\) or \(\overline{p_v}\).
In addition, at any marking at most one of such transitions is enabled because \(m(p_v) + m(\overline{p}_v) = 1\) holds.
Hence, for any update scheme in \(\mathcal{M}\), we have a corresponding firing scheme in \(\mathcal{P}\), which preserves the equivalence between the dynamics of \(\mathcal{M}\) and \(\mathcal{P}\)~\cite{DBLP:journals/nc/ChatainHKPT20}.

For illustration, let us reconsider the Boolean network shown in Example~\ref{example:BN}.
Figure~\ref{fig:PN} shows the Petri net encoding of this Boolean network.
Place \(p_{x_1}\) (resp. \(\overline{p_{x_1}}\)) in \(\mathcal{P}\) represents the activation (resp.\ the inactivation) of node \(x_1\) in \(\mathcal{M}\).
Marking \(\{p_{x_1}, \overline{p_{x_2}}\}\) in \(\mathcal{P}\) represents state 10 in \(\mathcal{M}\).
Transitions \(t^{1}_{x_1}\) and \(t^{2}_{x_1}\) represent the update of node \(x_1\).
Of course, in any marking \(t^{1}_{x_1}\) and \(t^{2}_{x_1}\) cannot be both enabled.
Then, the fully asynchronous update scheme in \(\mathcal{M}\) corresponds to the classical firing scheme in \(\mathcal{P}\) where only one of the enabled transitions for a given marking will be fired~\cite{Murata1989}.

Note that given a Boolean network in the standard SBML-Qual format~\cite{chaouiya2013sbml}, i.e., the package of SBML v3~\cite{keating2020sbml} for such models, one can easily obtain its Petri net encoding in the Petri Net Markup Language  (PNML)\footnote{\url{https://www.pnml.org/}} standard using the BioLQM\footnote{\url{http://www.colomoto.org/biolqm/}} library.
This piece of software extracted from GINsim~\cite{chaouiya2012logical} and part of the CoLoMoTo\footnote{\url{http://colomoto.org/}}~\cite{naldi2015cooperative} software suite allows for easy conversion between standard formats.
It also accepts many other common formats for Boolean networks, notably the \verb|.bnet| files of the  BoolNet~\cite{mussel2010boolnet,klarner2017pyboolnet} tools.
The conversion is executed as follows:

@@

Note that transforming a Boolean network defined by its functions into its Petri net encoding roughly relies on obtaining conditions for the activation and inactivation of the states. In~\cite{chaouiya2004qualitative} this took the form of the whole truth table of the Boolean functions, but as shown in Appendix 1 of~\cite{chatain2014characterization} computing Disjunctive Normal Forms (DNF) of each Boolean function is enough.
Though this might appear quite computationally intensive it is important to remark first that contrary to the prime-implicants case, there is no need to find \emph{minimal} DNFs.
One way to look at this is to consider that this amounts to a similar approach as that used in~\cite{DBLP:conf/ictai/ChevalierFPZ19} but with the encoding of both activation and inhibition functions as DNFs in order to take into account possible non-local-monotonicity.
This does not change the worst-case-complexity (obtaining a single DNF being exponential) but might matter a lot in practice.
As such, we will explore how this transformation, here using BDDs in BioLQM, and the one based on the most-permissive semantics compare in the Section~\ref{sec:eval} on evaluation.

@@

\section{Minimal trap spaces as maximal conflict-free siphons}
\label{sec:Main_finding}

First, we add a definition related to any set of places of a Petri net encoding a Boolean network, and notably a siphon of such a net.

@@

  A set of places of Petri net \(\mathcal{P}\) encoding Boolean network \(\mathcal{M}\) is \emph{conflict-free} if it does not contain any two places corresponding to the active and inactive states of the same \emph{node} of \(\mathcal{M}\).
  Then, a conflict-free siphon \(S\) is said to be \emph{maximal} if and only if there is no other conflict-free siphon \(S'\) such that \(S \subset S'\).

@@

Intuitively, a siphon is a set of places that once unmarked remains so.
If it is conflict-free then its dual corresponds to a partial-state of the model such that whatever update, the fixed values remain so (since the unmarked places remain unmarked).
This is precisely the definition of a trap space and maximality of the siphon is equivalent to as many fixed values as possible, hence minimality of the trap space.
For example, the Boolean network given in Example~\ref{example:BN} has two trap spaces, \(m_1 = 11\) and \(m_2 = \star\star\).
The Petri net encoding of this Boolean network has five generic siphons, \(S_1 = \emptyset\), \(S_2 = \{p_{x_1}, \overline{p_{x_1}}\}\), \(S_3 = \{p_{x_2}, \overline{p_{x_2}}\}\), \(S_4 = \{\overline{p_{x_1}}, \overline{p_{x_2}}\}\), and \(S_5 = \{p_{x_1}, \overline{p_{x_1}}, p_{x_2}, \overline{p_{x_2}}\}\).
However, only \(S_1\) and \(S_4\) are conflict-free siphons and correspond to \(m_2\) and \(m_1\), respectively.
Since \(S_1 \subset S_4\), \(S_4\) is a maximal siphon corresponding to the minimal trap space \(m_1\).
Hereafter, we formally prove that a maximal conflict-free siphon is equivalent to a minimal trap space.

@@

\subsection{Randomly generated models}
